What's hot and what's not:\\Windowed developer topic analysis

### 2009-04-11 lda-paper.tex
removed a bunch of comments
main edits and spelling

%   At different times, developers focus on different topics and tasks
%   during development.  These different topics and tasks might
%   be obvious to developers familiar with the project, but other
%   stake-holders, like managers, might be less aware of the topics of
%   development. Imagine the scenario where developers unexpectedly had to deal with
%   issues orthogonal to the features suggested, at the end of the
%   iteration a manager would wonder where all the effort went to.
%   Currently it is difficult for stake-holders to
%   audit what actually occurred and when certain features or topics
%   were focussed on. We propose a visualization of the
%   changing development topics over time. Our approach uses a topic
%   analysis tool like Latent Dirichlet Allocation (LDA) or Latent
%   Semantic Indexing (LSI). Instead of applying the tool to the
%   entire corpus we apply it to windows (periods) of the corpus. This allows us to
%   see the evolving stream of topics of development occurring
%   over time.

%   During different periods, developers focus on different topics and
%   different tasks during development. Stake-holders, such as managers,
%   not directly involved with the development might have less awareness
%   of the project and the topics and tasks that the developers are
%   focussed on.  Imagine the scenario where a manager is trying to
%   discover where all the developers' effort went after features have
%   not been implemented because the developers unexpectedly had to deal
%   with issues orthogonal to these features.  We propose to help
%   stake-holder understand the topics of development by extracting
%   topics from the descriptions of changes that programmers provide
%   when they make changes.  We leverage topic analysis tools such as
%   Latent Dirichlet Allocation (LDA) or Latent Semantic Indexing (LSI)
%   to analyze periods, such as months, within a project's
%   lifetime. Previous work applied LDA and LSI to
%   to a project's entire lifetime.  These tools associate
%   descriptions with independent topics extracted from these
%   descriptions.  We propose a visualization of changing development
%   topics over time that allows us to see the evolving stream of topics
%   of development occurring over time. We also show that windowed topic
%   analysis offers benefits compare to entire lifetime topic analysis
%   because of the independence of topics between windows.

% During different periods, developers focus on different topics and
% different tasks during development. Stake-holders, such as managers, who are
% not directly involved with the project's development might have less awareness of the
% project and the topics and tasks that the developers have focussed on.
% Imagine the scenario where a manager is trying to discover where all the
% developers' effort went after features have not been implemented because the
% developers unexpectedly had to deal with issues orthogonal to these
% features.  We propose to help stake-holders understand the topics of
% development by extracting topics from the descriptions of changes that
% programmers provide when they make commits.  We leverage topic analysis
% tools such as Latent Dirichlet Allocation (LDA) or Latent Semantic Indexing
% (LSI) to analyze periods, such as months, within a project's lifetime,
% whereas previous work applied LDA and LSI to to a project's entire lifetime.  These
% tools associate descriptions with independent topics extracted from these
% descriptions.  
% We propose a visualization of changing development topics
% over time that allows us to see the evolving stream of topics of development
% occurring over time. We demonstrate that windowed topic analysis offers
% benefits over entire lifetime topic analysis because of the
% independence of topics between windows.

@@

%   In this lab report we seek to explore how topics shift over time in
%   the source control repository of a project. We analyze commit
%   comments, and even source code per revision per each time window and
%   extract the topics prevalent in that window. We expect that topic
%   sets will change over time and the change in topic sets indicate a
%   change in the focus of development. We suppose this could help us to
%   segment the revisions on the time line to discover parts of an
%   iteration. For topic extraction we use \lda (LDA).

@@

%?

%

% In the presence of fine grained changes and detailed change-logs

% Developers topics the focus of develop

% many change-logs how to sort by change type

% interleaved changes relate similar changes

What happened in the previous maintenance or development iteration?
What were the developers working on? What was the developer in the
next cubicle working on?  Which common topics and issues were dealt
with in the current release of the software? What were the topics of
the previous release?  Given the requirements agreed to at the start
of the iteration did our developers work on them? What else did they
work on?

% XXX Come back here, clear up and cut 
% Topic analysis, with respect to
% Software Control Systems (SCSs) such as CVS, Subversion, Git etc.,
% is potentially valuable as it has many uses for various
% stakeholders.  Imagine a development team has agreed to implement
% some features at the start of a development iteration.  By the end
% of the iteration only some of the agreed upon features are
% implemented.  The developers insist they were focusing on the
% features agreed upon, but what actually happened?  Topic analysis
% could help by teasing out the distinct topics or focusses of the
% developers.  Perhaps the developers had to deal with other pressing
% issues.  Developers might have grouped those orthogonal issues into
% the features they were working on, thus when asked about how their
% time was spent the orthogonal changes would not be covered.  This is
% the kind of underlying story that we hope that windowed topic
% analysis can uncover. If windowed topic analysis can discover
% important but overlooked topics, it can help both developers and
% managers.  Developers and Managers can see a summary of what worked
% on and the topics of development.

@@

%It is apparent that we have to compare these topics some how by
%item-set or similarity.

% \begin{table*}
% \centering
% \begin{tabular}{|ccc|l|}
% \hline
% 2000 &  Jul &  31 &    chmod \\
% 2000 &  Sep &  29 &    fixes benchmark logging windows \\
% 2000 &  Nov &  28 &    typo fix insert\_multi\_value \\
% 2001 &  Jan &  27 &    fixes Innobase Cleanups auto-union \\
% 2001 &  Mar &  28 &    2 topics bugfix, logging , TEMPORARY,  \\
% \hline
% 2001 &  Jul &  26 &    update Allow TABLES LOCK [a] \\

% 2001 &  Aug &  25 &    tables row version [a] \\
% \hline
% 2001 &  Sep &  24 &    update checksum merge \\
% 2001 &  Oct &  24 &    fixed fix \\
% 2001 &  Dec &  23 &    HPUX SCO fix \\
% \hline
% 2002 &  Feb &  21 &    net buffer length  max\_allowed\_packet [b] \\
% 2002 &  Mar &  23 &    small buf fix [b]  \\
% \hline
% 2002 &  May &  22 &    [popular] fix SCO OSF1 table\_name \\
% 2002 &  Nov &  18 &    HPUX11 compiler HP \\
% \hline
% 2003 &  Feb &  16 &    Linux errno  [c] \\
% 2003 &  Mar &  18 &    alarm bookmark bug [c] \\
% \hline
% 2003 &  Sep &  14 &    Auto logging merge windows distribution fix 64-bit 4.0 Cleanup \\
% \hline
% \end{tabular}
% \caption{Tracking topics associated with the word portability, note some continuous blocks}
% \label{tab:portability}
% \end{table*}

@@

%\shrinkit
%\subsection{Datasets and Tools}
%\shrinkit

%   1. [ ] LDA

%   2. [ ] Extractor

%   3.   Similarity

%   4.   Smearing

%   5.   Visualizer

%LDA Skeleton
%1. [ ] Abstract 
%2. [ ] Introduction

%3. [ ] Previous Work [0/2]
%   2. [ ] LDA Work [0/6]

@@

% Extractors:
% \begin{itemize}
% \item CVSSuck - CVS Suck mirrors RCS files from a CVS repository. 
% \item softChange - extract CVS facts to a PostgreSQL database.
% \item bt2csv - Convert BitKeeper repositories to facts in CSV
%   databases.
% \end{itemize}

%   1. Interesting smears 
%   2.   postgresql

@@

%XXX Redo?

@@

% When we examined PostgreSQL, we did not find many trends with two or
% more topics; however, we did find one large trend which stretched
% across most of its existence.  Much of the large trend consisted of
% stop words but it did include topics which had non-stop words such as
% \emph{patch}, \emph{fix}, \emph{update}.

% The second largest topic was interesting because it concerned changes from
% 1998, from developers using Central Europe Time (CEST), where the two
% notable tokens were 1998 and CEST. These changes appear to be patches
% submitted by one particular developer who lived in that timezone.

% The third largest trend we found appeared to concerned with copying
% and moving files between branches as it referenced the old branch.
% branches.  We inferred this trend based on the fact that the old
% version identifier was referenced, and we know this this is common
% practice when performing a branch merge, since CVS cannot record such
% information directly.

% Once stop words were removed the some of the purposes became more
% clear. The large trend disappeared, the largest trend became bug fixes
% submitted by a developer named Dal Zotto which is a short trend.

%XXX REDO POSTEGRESQL?
% Large Trends is not the point!!!
% That's what normal LDA gives us

%But large trends are not the point, by investigating windowed trends
%we want more local information about trends being worked on.

@@

%XXX REDO?

@@

% We tracked Firebird from August 2000 to January 2006.
% Our first attempt with Firebird resulted in a set of trends that were
% dominated by one large trend containing tokens that were mostly stop
% words such as references to one author
% \texttt{carlosga05} and the words: \emph{the, to, a, for, of, is, it}.

% The second largest trend related to branching and possibly merging; it
% contained language such as ``was added on branch'' and contained
% references to tags and branches.

% The third largest trend contained references to the changelog, the
% build system, architecture and platform support. It specifically
% mentioned tokens such as \emph{Solaris}, \emph{port} and \emph{x86}.

% Other trends included topics regarding AIX PPC compilation, updating
% the build process, internationalization and UTF8, Darwin build support
% and bug fixing.

% Topics that were not trends but appeared to be  interesting were mostly external bug
% fixes submitted to the project.
%  In these cases, the developers would express thanks in their commit log comments, such as ``Thanks, Bill Lam''.  Other easily discernible topics included 
%  tokens and topics such as: \emph{compiler workarounds}, \emph{nightly updates}, \emph{packets} and \emph{MSVC++}.

@@

%XXX REDO?

%XXX Migod todo
% I think that you should enlarge this discussion of MaxDB (at the
% expense of the other systems) and make careful references to how the
% diagrams give you extra info about the evolution of the system (in
% terms of topics).

@@

The plots we produced of MaxDB 7.500 were unlike those of the other
systems, as there was a period where no real development occurred and
thus there were no topics or trends whatsoever (see the gap in Figure
\ref{fig:topicsmear}). Using a topic similarity of $7/10$ we evaluated
of MaxDB 7.500, the first period from June 2004 to January 2005, and
the second period from June 2005 to June 2006.

The largest common trend has references to build system files like
\texttt{SYSDD.punix} and \texttt{MONITOR.punix}.  This trend is
partially displayed at the top of the zoomed in compact-trend-view
(Figure \ref{fig:zoomedsmear}) and at the top of the trend-histogram
(Figure \ref{fig:histogram}).  Other tokens mentioned are
\emph{Sutcheck v1.09} (the prefix SUT stands for Storable Unit
Type), a tool that would also automate check-ins using a Perforce SCS tool
which was exporting check-ins to CVS.

@@

The third most common trend, seen on Figure \ref{fig:trendtimeline},
seemed to include tokens related to operating system support, such as
Linux and Windows, as well as architecture support, AMD64 and
Opteron. The word \emph{problem} was common among all of these
trends. This trend seemed related to the smaller fourth largest trend
that had tokens \emph{AMD64} and \emph{Windows}. This shows that
topics can overlap but still not match via our similarity measure.

%XXX Edit
% Was reading lda.report
Bug tracker URLs dominated unique topics during some months. For
instance in the last month of MaxDB 7.500 development every topic
contained 1 unique Bug tracker URL, this pattern did not occur in the
previous month. We investigated the actual revisions and we did find
there was difference in behavior between, and that the last month of
development commits were mostly referencing the bug tracker. If the
topics of one month were 20 or more unique bug tickets being
addressed, the global LDA analysis would probably miss this, yet these
bug tickets were the focus of development for that month and not
necessarily globally relevant.

The query optimizer was a topic that recurred during MaxDB's
development. In our plots, topics that mention \emph{optimizer} occur four
times, yet in the global-trend-view (Figure \ref{fig:statictopics}) it
is not in any of the topics. A query optimizer is rather important to
an DBMS, but as we have shown it just does not pop up as a topic on
its own. We tried to remove words to see if we could get an 
\emph{optimizer} topic. After removing stop words and then two of the most
common words, the global analysis finally had a topic with optimizer in
its topic ten words. This shows that optimizer was important but it
had been obscured, while it would have been noticed using a more
locally relevant approach that we propose.

We noticed that commits that mentioned \emph{PHP} occurred two thirds
less frequently than commits that mentioned \emph{Perl}, but
\emph{Perl}-related topics appeared in the global static topics for
MaxDB while \emph{PHP}-related topics did not.  Our local topic
analysis mentioned \emph{PHP} in 5 different topics, yet only
mentioned \emph{Perl} in 4 different windowed topics and 1 global
topic. Perhaps this is because there was a cluster of \emph{Perl}
mentions during one month while \emph{PHP} was mentions were more
spread out.

We noticed that \emph{Perforce} and \emph{Changelist} were in just
about every topic, so we added them to the stop words list. As a
result, longer trends were shortened and in some cases the total
number of topics found per month was less.  Evaluating different
similarity thresholds showed that by removing common words one reduces
the general similarity of topics. That said, the larger topics still
clearly apparent. Thus if more relevant stop words are added one
should tune the topic similarity parameters to handle these changes.

% UNICODE

% DEC/OSF1

% ODBC

% crash

% PHP yet Perl is mentioned is less but it part

% SQLCLI
% %V75

% SYSDD.cpnix
% backup

% select
% make
% memory
% view

% debug

% V75 is mentioned many times maxdb version guh

@@

%XXX Migod notes This section is perhaps the most important of the
% whole paper.  You need expand it.  Maybe you should move the first
% two paras of the conclusions in here too.

@@

Previous work on topic analysis that employed LSI and LDA typically
extracted a specified number of topics and then tracked them over time. This
means that a single, static set of topics was tracked across the whole
development history of a project. \emph{Global topic analysis} is
topic analysis applied to entire lifetime of a project.

@@

If we approach the difference of global topic analysis and windowed
topic analysis via common tokens we can see that common token tend to
dominate both results, at least in the case of MaxDB 7.500 our local
topic approach netted us topics which contained these relatively
important and common words: \emph{UNICODE}, \emph{DEC/OSF1},
\emph{ODBC}, \emph{crash}, \emph{SQLCLI}, \emph{SYSDD.cpnix},
\emph{backup}, \emph{select}, \emph{make}, \emph{memory}, \emph{view},
and finally \emph{debug}. \emph{ODBC} is quite important because it
often determines how databases communicate with software. None of
these tokens were part of the global topic analysis topics, but they
were part of 566 commits (6\% of the entire system) to MaxDB
7.500. These tokens were part of 87 out of 520 (26 months, 20 topics
per month) of our locally generated topics.

@@

The number of commits per month is inconsistent as some months have a
lot of changes while others months have almost none.
%   1.   validation
%   2.   LDA is questionable
%   3.   blackbox problem
%   4.   not in the token
%   5.   multiple X token
%9.    Future Work

@@

% Even with our liberal topic similarity metrics that produced both
% long and short trends, we showed that there are only a few trends in a
% repository which recur. Since so few trends recur and so many
% trends appear only once this suggests that global topic analysis might
% be ignoring locally unique topics.

% The utility of global topic analysis is questionable if the value of
% information decreases as it becomes older. Perhaps older trends will
% dominate the topic analysis. Windowed localized topic analysis show
% what are the unique topics.

We proposed and demonstrated the application of windowed topic
analysis, that is, topic analysis applied to commit messages during
periods of development. This approach differs from previous work that
would applied topic analysis globally to the entire history of a
project without regard to time.. We showed that many topics that exist
locally are relevant and interesting yet would often not be detected
via global topic analysis. We identify recurring topics with a topic
similarity measure which looks for topics which recur and mutate
repeatedly throughout the development of the software project.

The value of windowed topic analysis was demonstrated on a case study
of three database systems with a particular focus on MaxDB 7.500. We
showed that global topic analysis missed important topics like such as
ODBC changes.

%XXX summarize our position

%XXX Add more detail here

We presented several visualization techniques that focused on
different aspects of the data: temporality of trends, trend size, and
a compact-trend-view. The compact-trend-view shows much more
information than the views that global analysis could show, it
indicates how focused a period is by the total number of topics as
well shows topics by similarity so one can track trends across time.
Our trend-histogram immediately hi-lights and measures the size of
trends while our trend-time-line view shows how a topic reoccurs over
time. These visualizations help us get an over feeling about the
common and unique topics that developers focus on. If implemented
properly a user can easily zoom and query a topic.

>>>>

@@

@@

%?
What happened in the previous maintenance or development iteration?
What were the developers working on? What was the developer in the
next cubicle working on?  Which common topics and issues were dealt
with in the current release of the software? What were the topics of
the previous release?  Given the requirements agreed to at the start
of the iteration did our developers work on them? What else did they
work on?

@@

@@

@@

@@

@@

@@

@@

@@

% Migod todo
% I think that you should enlarge this discussion of MaxDB (at the
% expense of the other systems) and make careful references to how the
% diagrams give you extra info about the evolution of the system (in
% terms of topics).

@@

The plots we produced of MaxDB 7.500 were unlike those of the other
systems, as there was a period where no real development occurred and
thus there were no topics or trends whatsoever (see the gap in Figure
\ref{fig:topicsmear}). Using a topic similarity of $7/10$ we evaluated
MaxDB 7.500. MaxDB 7.500's first period was from June 2004 to January
2005, and its second period was from June 2005 to June 2006.

The largest common trend has references to build system files like
\texttt{SYSDD.punix} and \texttt{MONITOR.punix}.  This trend is
partially displayed at the top of the zoomed in compact-trend-view
(Figure \ref{fig:zoomedsmear}) and at the top of the trend-histogram
(Figure \ref{fig:histogram}).  Other tokens mentioned are
\emph{Sutcheck v1.09} (the prefix SUT stands for Storable Unit Type),
that is a tool that would also automate check-ins using a Perforce SCS
tool which was exporting check-ins to CVS.

@@

The third most common trend, seen on Figure \ref{fig:trendtimeline},
seemed to include tokens related to operating system support, such as
\emph{Linux} and \emph{Windows}, as well as architecture support, \emph{AMD64} and
\emph{Opteron}. The word \emph{problem} was common among all of these
trends. This trend seemed related to the smaller fourth largest trend
that had tokens \emph{AMD64} and \emph{Windows}. This shows that
topics can overlap but still not match via our similarity measure.

%XXX Edit
% Was reading lda.report
Bug tracker URLs dominated unique topics during some months. For
instance in the last month of MaxDB 7.500 development every topic
contained 1 unique Bug tracker URL, this pattern did not occur in the
previous month. We investigated the actual revisions and we did find
there was a difference in behavior between, and that the last month of
development commits were mostly referencing the bug tracker. If the
topics of one month were about unique bug tickets being
addressed, the global topic analysis would probably miss this, yet these
bug tickets were the focus of development for that month but not
necessarily globally relevant.

The query optimizer was a topic that recurred during MaxDB's
development. In our plots, topics that mention \emph{optimizer} occur four
times, yet in the global-trend-view (Figure \ref{fig:statictopics}) it
is not in any of the topics. A query optimizer is rather important to
an DBMS, but as we have shown it just does not pop up as a topic on
its own. We tried to remove words to see if we could get an 
\emph{optimizer} topic. After removing stop words and then two of the most
common words, the global analysis finally had a topic with optimizer in
its topic ten words. This shows that optimizer was important but it
had been obscured, while it would have been noticed using the more
locally relevant approach that we propose.

We noticed that commits that mentioned \emph{PHP} occurred two thirds
less frequently than commits that mentioned \emph{Perl}, but
\emph{Perl}-related topics appeared in the global static topics for
MaxDB while \emph{PHP}-related topics did not.  Our local topic
analysis mentioned \emph{PHP} in 5 different topics, yet only
mentioned \emph{Perl} in 4 different windowed topics and 1 global
topic. Perhaps this is because there was a cluster of \emph{Perl}
mentions during one month while the \emph{PHP} mentions were more
spread out.

We noticed that \emph{Perforce} and \emph{Changelist} were in just
about every topic, so we added them to the stop words list. As a
result, longer trends were shortened and in some cases the total
number of topics found per month was less.  Evaluating different
similarity thresholds showed that by removing common words one reduces
the general similarity of topics. That said, the larger topics were still
clearly apparent. Thus if more relevant stop words are added one
should tune the topic similarity parameters to handle these changes.

% UNICODE
% DEC/OSF1
% ODBC
% crash
% PHP yet Perl is mentioned is less but it part
% SQLCLI
% %V75
% SYSDD.cpnix
% backup
% select
% make
% memory
% view
% debug
% V75 is mentioned many times maxdb version guh

@@

% Migod notes This section is perhaps the most important of the
% whole paper.  You need expand it.  Maybe you should move the first
% two paras of the conclusions in here too.

@@

Previous work on topic analysis that employed LSI and LDA typically
extracted a specified number of topics and then tracked them over
time. This means that a single, static set of topics was tracked
across the whole development history of a project. \emph{Global topic
  analysis} is topic analysis applied against the entire lifetime of a
project.

@@

If we approach the difference of global topic analysis and windowed
topic analysis via common tokens we can see that common tokens tend to
dominate both results, at least in the case of MaxDB 7.500 our local
topic approach netted us topics which contained these relatively
important and common words: \emph{UNICODE}, \emph{DEC/OSF1},
\emph{ODBC}, \emph{crash}, \emph{SQLCLI}, \emph{SYSDD.cpnix},
\emph{backup}, \emph{select}, \emph{make}, \emph{memory}, \emph{view},
and finally \emph{debug}. \emph{ODBC} is quite important because it
often determines how databases communicate with software. None of
these tokens were part of the global topic analysis topics, but they
were part of 566 commits (6\% of the entire system) to MaxDB
7.500. These tokens were part of 87 out of 520 (26 months, 20 topics
per month) of our locally generated topics.

@@

The number of commits per month is inconsistent as some months have a
lot of changes while other months have almost none.
%   1.   validation
%   2.   LDA is questionable
%   3.   blackbox problem
%   4.   not in the token
%   5.   multiple X token
%9.    Future Work

@@

We proposed and demonstrated the application of windowed topic
analysis, that is, topic analysis applied to commit messages during
periods of development. This approach differs from previous work that
applied topic analysis globally to the entire history of a
project without regard to time. We showed that many topics that exist
locally are relevant and interesting yet would often not be detected
via global topic analysis. We identify recurring topics with a topic
similarity measure which looks for topics which recur and mutate
repeatedly throughout the development of the software project.

The value of windowed topic analysis was demonstrated on a case study
of three database systems with a particular focus on MaxDB 7.500. We
showed that global topic analysis missed important topics such as
\emph{ODBC}.

We presented several visualization techniques that focused on
different aspects of the data: temporality of trends, trend size, and
a compact-trend-view. The compact-trend-view shows much more
information than the views that global analysis could show, it
indicates how focused a period is by the total number of topics, as
well, it shows topics by similarity so one can track trends across
time.  Our trend-histogram immediately highlights and measures the
size of trends while our trend-time-line view shows how a topic
reoccurs over time. These visualizations help us get an overall
feeling about the common and unique topics that developers focus
on. If implemented properly a user could easily zoom in and query a
summary of a topic or trend.
